Financial Contributions to Presidential Campaign by SOHAN SAMANTA
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(reshape)
library(reshape2)
library(dplyr)
library(maps)
library(GGally)
library(scales)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(lubridate)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
IL <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/IL2.csv', 
               header = TRUE, row.names = NULL)
row.names <- names(IL)[2:length(names(IL))]
names(IL) <- row.names
IL <- IL[, 1:18]
# changing date and time format
IL$contb_receipt_dt <- dmy(IL$contb_receipt_dt)
head(IL)

# special operator
'%!in%' = Negate('%in%')
```

> **Tip**: Before you create any plots, it is a good idea to provide a short
introduction into the dataset that you are planning to explore. Replace this
quoted text with that general information!

UPDATE THE DATA - insert party details
```{r}
republican <- c("Trump, Donald J.", "Cruz, Rafael Edward 'Ted'", "Walker, Scott", 
                "Bush, Jeb", "Rubio, Marco", "Christie, Christopher J.", "Kasich, John R.",
                "Paul, Rand", "Graham, Lindsey O.", "Fiorina, Carly", "Jindal, Bobby",
                "Santorum, Richard J.", "Huckabee, Mike", "Carson, Benjamin S.", 
                "Perry, James R. (Rick)", "Pataki, George E.")
democratic <- c("Clinton, Hillary Rodham", "Sanders, Bernard", "O'Malley, Martin Joseph",
                "Lessig, Lawrence", "Webb, James Henry Jr.")
libertarian <- c("Johnson, Gary")
green <- c("Stein, Jill")
independent <- c("McMullin, Evan")

IL$party = ""
IL$party[IL$cand_nm %in% republican] = 'republican'
IL$party[IL$cand_nm %in% democratic] = 'democratic'
IL$party[IL$cand_nm %in% green] = 'green'
IL$party[IL$cand_nm %in% independent] = 'independent'
IL$party[IL$cand_nm %in% libertarian] = 'libertarian'
```

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
# grouping the data according to candidates and party
IL.cand_party <- IL %>%
  group_by(cand_nm, party) %>%
  summarise(receipt_amt = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(party)

# to order the members according to the money they received
IL.cand_party$cand_nm <- with(IL.cand_party, factor(cand_nm, levels = cand_nm[order(receipt_amt)]))


ggplot(aes(x = cand_nm, y = receipt_amt), data = subset(IL.cand_party))+
  geom_bar(stat = 'identity', position = 'dodge', aes(fill = party))+ 
  coord_flip()
```

```{r}
IL.party <- IL %>%
  group_by(party) %>%
  summarise(receipt_amt = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(party)

ggplot(aes(x = party, y = receipt_amt), data = IL.party)+
  geom_bar(stat = 'identity', aes(fill = party))
```

So, in Illinois, the democrats get more donations than any other party. Almost 2 times more. And even then, Hillary gets a major share of the donations that the whole party receives. So let us take a quick look at the percentage of donations the members receive.

```{r}
# Reshaping IL.cand_party
IL.cand_party.wide <- dcast(IL.cand_party,
                          party ~ cand_nm,
                          value.var = 'receipt_amt')

for(i in 1:nrow(IL.cand_party.wide)){
  IL.cand_party.wide$total[i] <- sum(IL.cand_party.wide[i,2:25], na.rm = TRUE)
}

IL.cand_party.long <- melt(IL.cand_party.wide,
                          id.vars = c('party', 'total'))

IL.cand_party.long <- subset(IL.cand_party.long, !is.na(value))
names(IL.cand_party.long) <- c('party','total','cand_nm','receipt_amt')

IL.cand_party.long$percent <- with(IL.cand_party.long, (receipt_amt*100/total))

# lets look at the percentage of only democrats and republicans

p1 <- ggplot(aes(x = cand_nm, y = percent), 
             data = subset(IL.cand_party.long, party == 'democratic'))+
  geom_histogram(stat = 'identity', aes(fill = I('red4')))+
  ggtitle('Democratic')+
  labs(y = '% of contribution received')+
  theme(axis.text.x = element_text(angle = 90), axis.title = element_text(size = 7),
        axis.title.y = element_text(size = rel(2)), axis.title.x = element_text(size = rel(2)))

p2 <- ggplot(aes(x = cand_nm, y = percent), 
             data = subset(IL.cand_party.long, party == 'republican'))+
  geom_histogram(stat = 'identity', aes(fill = I('dodgerblue4')))+
  ggtitle('Republican')+
  labs(y = '% of contribution received')+
  theme(axis.text.x = element_text(angle = 90), axis.title = element_text(size = 7),
        axis.title.y = element_text(size = rel(2)), axis.title.x = element_text(size = rel(2)))

grid.arrange(p1, p2, ncol = 2)

```


lets look at the individual contributions, for the main receipeints

```{r}
# First lets group the data according to contributors, candidates and the parties
IL.contributor <- IL %>%
  group_by(contbr_nm, cand_nm, party) %>%
  summarise(total_contribution = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(contbr_nm)

# Now lets look at one party at a time
ggplot(aes(x = cand_nm, y = total_contribution), 
       data = subset(IL.contributor, party == 'democratic'))+
         geom_boxplot()+
  scale_y_continuous(limits = c(0,10000), breaks = seq(0,10000,500))

summary(subset(IL.contributor, party == 'democratic'))  
```

#republican party
```{r}
ggplot(aes(x = cand_nm, y = total_contribution), 
       data = subset(IL.contributor, party == 'republican'))+
         geom_boxplot()+
  scale_y_continuous(limits = c(0,20000), breaks = seq(0,20000,500))+
  theme(axis.text.x = element_text(angle = 90))

summary(subset(IL.contributor, party == 'republican')) 
```


Donald Trump and Bernie Snaders have donations of the same range. How do their summaries stack against each other
```{r}
summary(subset(IL.contributor, cand_nm == 'Trump, Donald J.'))  
summary(subset(IL.contributor, cand_nm == 'Sanders, Bernard'))  
```
Freqpoly of above data
```{r}
# A grouping to be primarily used in the later stage, but brought forward for this plot
IL.party_date <- IL %>%
  group_by(contb_receipt_dt, cand_nm, party) %>%
  summarise(total_amnt = sum(contb_receipt_amt),
            mean_amnt = mean(contb_receipt_amt),
            median_amnt = median(contb_receipt_amt),
            n = n()) %>%
  arrange(contb_receipt_dt)

# Freqpoly
ggplot(aes(x = total_amnt), 
       data = subset(IL.party_date, cand_nm %in% c('Trump, Donald J.','Sanders, Bernard')))+
  geom_freqpoly(aes(color = cand_nm))+
  scale_colour_brewer(type = 'qual', guide = guide_legend(title = 'candidate', reverse = F))+
  xlim(c(0,NA))+
  ylim(c(0,NA))
```

Compare the donors on the basis of their donation amounts
```{r}
# what amount does the donors donate. Summary of doation
summary(IL.contributor$total_contribution)

ggplot(aes(x = total_contribution), data = IL.contributor)+
  geom_histogram(binwidth = 10, color = I('black'), aes(fill = I('Orange')))+
  scale_x_continuous(limits = c(0,1500), breaks = seq(0, 1500, 50))
```

```{r}
IL.contributor$party <- with(IL.contributor, factor(party, 
                        levels = c('democratic','republican','libertarian',
                                       'green','independent')))
str(IL.contributor$party)


ggplot(aes(x = total_contribution, y = n, color = party), data = IL.contributor)+
  geom_point(alpha = 0.4, position = 'jitter')+
  xlim(0, quantile(IL.contributor$total_contribution,0.9))+
  ylim(0, quantile(IL.contributor$n,0.9))+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'party', reverse = F))
```

Now let us look at the contributions through dates
```{r}
ggplot(aes(x = contb_receipt_dt, y = total_amnt, color = party), 
       data = IL.party_date)+
  geom_line(na.rm = TRUE)+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'party', reverse = F))+
  scale_x_date(limits = c(dmy('01-01-2015'), dmy('01-01-2017')))+
  ylim(c(0,NA))
```

each party, and individual candidates
```{r}
p1 <- ggplot(aes(x = contb_receipt_dt, y = total_amnt, color = cand_nm), 
       data = subset(IL.party_date, party == 'democratic'))+
  geom_line(na.rm = TRUE)+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'candidate', reverse = F))+
  scale_x_date(limits = c(dmy('01-01-2015'), dmy('01-01-2017')))+
  ylim(c(0,NA))

exclude = c("Jindal, Bobby", "Santorum, Richard J.", "Huckabee, Mike", 
            "Perry, James R. (Rick)", "Pataki, George E.")

p2 <- ggplot(aes(x = contb_receipt_dt, y = total_amnt, color = cand_nm), 
       data = subset(IL.party_date, (party == 'republican' & cand_nm %!in% exclude)))+
  geom_line(na.rm = TRUE)+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'candidate', reverse = F))+
  scale_x_date(limits = c(dmy('01-01-2015'), dmy('01-01-2017')))+
  ylim(c(0,NA))

grid.arrange(p1, p2, ncol = 1)
```

Another interesting plot would be to look at the median amounts received by each party
```{r}
ggplot(aes(x = median_amnt), data = IL.party_date)+
  geom_freqpoly(aes(color = party), na.rm = TRUE)+
  scale_colour_brewer(type = 'qual', guide = guide_legend(title = 'party', reverse = F))+
  xlim(c(0,NA))+
  ylim(c(0,175))
```

```{r}
# Read a new file that contained zip with corresponding longitudes and latitudes

## The following code to get the longitudes and latitudes take a long time. So I have done the 
## conversion once and saved the result in a file for easy access the next time I run the code

#zip <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/zip_codes_states.csv')

#IL.zip <- IL[c(3,5,7,10,11,19)]
#IL.zip$long = NA
#IL.zip$lat = NA

# inserting long & lat
#for(i in 1:length(IL.zip$contbr_zip)){
#   indices <- which(zip$zip_code == IL.zip$contbr_zip[i])
#   IL.zip$lat[i] <- ifelse(length(indices)>0,zip$latitude[indices],NA)
#   IL.zip$long[i] <- ifelse(length(indices)>0,zip$longitude[indices],NA)
#}

#write.csv(IL.zip, file = "IL_zip.csv")

IL.zip <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/IL_zip.csv')
```

```{r}
all_states <- map_data("state")

# state map with dots
ggplot(aes(x = long, y = lat), data = IL.zip) +
  geom_polygon(data = all_states, aes(x=long, y=lat, group = group),colour="white", fill="grey10")+
  geom_point(color = "dark red", na.rm = TRUE)
```


```{r}
# only chicago with dots
ggplot(aes(x = long, y = lat), data = subset(IL.zip, long >= '-87.49638' & long <= '-91.50136'
                                             & lat >= '37.00161' & lat <= '42.50774')) +
  geom_polygon(data = subset(all_states, region == 'illinois'), 
               aes(x=long, y=lat, group = group),colour="white", fill="grey10")+
  geom_point(color = "dark red", na.rm = TRUE, alpha = 0.4, position = 'jitter')
```

# On a map
```{r}
# scaling dot size
IL.zip_grouped <- subset(IL.zip, contb_receipt_amt > 0)

IL.zip_grouped$amnt[IL.zip_grouped$contb_receipt_amt >= 80] <- '10800'
IL.zip_grouped$amnt[IL.zip_grouped$contb_receipt_amt >= 27 & 
                      IL.zip_grouped$contb_receipt_amt <= 80] <- '80'
IL.zip_grouped$amnt[IL.zip_grouped$contb_receipt_amt >= 15 & 
                      IL.zip_grouped$contb_receipt_amt <= 27] <- '27'
IL.zip_grouped$amnt[IL.zip_grouped$contb_receipt_amt <= 15] <- '15'

IL.zip_grouped$amnt <- as.numeric(IL.zip_grouped$amnt)

ggplot(aes(x = long, y = lat), data = subset(IL.zip_grouped, long >= '-87.49638' & long <= '-91.50136'
                        & lat >= '37.00161' & lat <= '42.50774' & contb_receipt_amt > 0)) +
  geom_polygon(data = subset(all_states, region == 'illinois'), 
               aes(x=long, y=lat, group = group),colour="white", fill="grey10")+
  geom_point(na.rm = TRUE, alpha = 0.5, position = 'jitter',
             aes(color = party, size = amnt))+
  scale_size_continuous(range=c(0.5,4.5), name = 'Amount')+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'party', reverse = F))
```

# With time
```{r}
IL.zip_grouped$date <- with(IL.zip_grouped, as.character(year(contb_receipt_dt)))

colors <- c("gold",
            "red4",
            "cornsilk",
            "darkgreen")

ggplot(aes(x = long, y = lat), data = subset(IL.zip_grouped, 
                                  long >= '-87.49638' & long <= '-91.50136'
                                  & lat >= '37.00161' & lat <= '42.50774' 
                                  & contb_receipt_amt > 0)) +
  geom_polygon(data = subset(all_states, region == 'illinois'), 
               aes(x=long, y=lat, group = group),colour="white", fill="grey10")+
  geom_point(na.rm = TRUE, alpha = 0.5, position = 'jitter',
             aes(color = date), shape = 21)+
  scale_color_manual(values = colors)
```

# I have zip wise money. I will place the county description on there.
# Then group by county. Then get the county wise population in there.
# So we get contributors from counties as well as the population of those counties
# then plot number of contributors vs population
```{r, message = FALSE}
IL.county_grouped <- IL.zip_grouped %>%
  group_by(contbr_zip, party) %>%
  summarise(amount = sum(contb_receipt_amt),
            n = n())

county_zip <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/county_zip.csv')  
county_zip$county <- as.character(county_zip$county)

IL.county_grouped$county <- ""
for(i in 1:length(IL.county_grouped$contbr_zip)){
  index <- which(county_zip$zip == IL.county_grouped$contbr_zip[i])
  IL.county_grouped$county[i] <- ifelse(length(index)>0, county_zip$county[index], NA)
}

county_zip.county <- IL.county_grouped %>%
  group_by(county, party) %>%
  summarise(amnt = sum(amount),
            num = sum(n))


county_population <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/illinois_county_population2.csv')
county_zip.county$population <- ""

for(i in 1:length(county_zip.county$county)){
  index <- which(county_population$County == county_zip.county$county[i])
  county_zip.county$population[i] <- ifelse(length(index)>0, county_population$Population[index], NA)
}

county_zip.county$population <- as.integer(county_zip.county$population)

ggplot(aes(x = population, y = num, color = party), data = county_zip.county)+
  geom_smooth(na.rm = TRUE)+
  scale_x_log10()+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'candidate', reverse = F))

```

With respect to occupation(total variation 8364, we wont look at them all)
```{r}
length(unique(IL$contbr_occupation))

occupation <- IL %>%
  group_by(contbr_occupation, party) %>%
  summarise(amount = sum(contb_receipt_amt),
            n=n()) %>%
  arrange(n)


occp = c('RETIRED','NOT EMPLOYED','ATTORNEY','TEACHER', 'PROFESSOR', 'PHYSICIAN', 
         'CONSULTANT','HOMEMAKER', 'LAWYER', 'SALES', 'STUDENT', 'ENGINEER')


ggplot(aes(x = party, y = amount), data = subset(occupation, contbr_occupation %in% occp))+
         geom_bar(stat = 'identity', aes(fill = party))+
         facet_wrap(~contbr_occupation)+
  theme(axis.text.x = element_text(angle = 45))
```

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!