Financial Contributions to Presidential Campaign by SOHAN SAMANTA
================================================================================

## INTRODUCTION
For this Exploratory Data Analysis project, I have decided to work with one of 
the Udacity suggested data sets dealing with the financial contributions to the
presidential election of 2016. The data sets have been divided by states, and 
here I have chosen to look at the state of Illinois.
The choice of Illinois is totally random, and the primary motivation behind that
is to start from neutral ground without any prior expectations of certain outcomes
or conclusions. As data scientists, we are supposed to be able to draw inferences 
from available data, not the other way round.
I start this project with a few initial questions in my mind. And as I progress 
I try to answer these:

  1. Contributions to parties and candidates are a direct representation of the 
  peoples support. So who garners the most support? Which party is in the lead, 
  and which candidates from individual parties are the best contenders? 
  
  2. What demographical conclusions can we deduce about the donors from the donations?
  
  3. What geographical aspects do these donations show?  Is there any relation
  between the population of a place and the number of donations?
  
  4. How do the contributions vary over time?
  
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# LIBRARIES
library(ggplot2)
library(reshape)
library(reshape2)
library(dplyr)
library(maps)
library(GGally)
library(scales)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(lubridate)
```

## THE DATASET
Let us first take a look at some of the data. 

```{r echo=FALSE, Load_the_Data}
# Load the Data
IL <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/IL2.csv', 
               header = TRUE, row.names = NULL)
row.names <- names(IL)[2:length(names(IL))]
names(IL) <- row.names
IL <- IL[, 1:18]
# changing date and time format
IL$contb_receipt_dt <- dmy(IL$contb_receipt_dt)
str(IL)

# special operator
'%!in%' = Negate('%in%')
```

## UPDATE THE DATA

We at once notice a few missing pieces that are required to answer the questions 
we set out to find. One primary missing information in the data set is the political 
party details, i.e. we do not get any information as to which candidate is affiliated
to which party. There are other missing links like the coordinates of the contributor,
the county population information, etc. But we will look at them at a later stage.

For now, let us insert the information into our data set and try to summarise the 
data.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
republican <- c("Trump, Donald J.", "Cruz, Rafael Edward 'Ted'", "Walker, Scott", 
                "Bush, Jeb", "Rubio, Marco", "Christie, Christopher J.", "Kasich, John R.",
                "Paul, Rand", "Graham, Lindsey O.", "Fiorina, Carly", "Jindal, Bobby",
                "Santorum, Richard J.", "Huckabee, Mike", "Carson, Benjamin S.", 
                "Perry, James R. (Rick)", "Pataki, George E.")
democratic <- c("Clinton, Hillary Rodham", "Sanders, Bernard", "O'Malley, Martin Joseph",
                "Lessig, Lawrence", "Webb, James Henry Jr.")
libertarian <- c("Johnson, Gary")
green <- c("Stein, Jill")
independent <- c("McMullin, Evan")

IL$party = ""
IL$party[IL$cand_nm %in% republican] = 'republican'
IL$party[IL$cand_nm %in% democratic] = 'democratic'
IL$party[IL$cand_nm %in% green] = 'green'
IL$party[IL$cand_nm %in% independent] = 'independent'
IL$party[IL$cand_nm %in% libertarian] = 'libertarian'

head(IL)
```

## SUMMARY OF THE DATA SET
```{r}
names(IL)
```

So we have 19 variables and 250,411 observations. Thats a lot of data!

```{r}
unique(IL$cand_nm)
unique(IL$party)
```
We have 24 candidates from the state of Illinois representing 5 political parties.
Next we see the contributor name, employment and their contributions. We also have 
the sip code and city name of the individual contributors.These are the points 
of information that at first glance jumps at me that I can start my work with and
get some idea as to their correlations.

## Univariate Plots Section
Lets start with a quick look at the cumulative contributions received by the 
individual candidates. We would also like some way to distinguish the candidates
by their parties, so we are color coding them.

```{r echo=FALSE, Univariate_Plots}
# grouping the data according to candidates and party
IL.cand_party <- IL %>%
  group_by(cand_nm, party) %>%
  summarise(receipt_amt = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(party)

# to order the members according to the money they received
IL.cand_party$cand_nm <- with(IL.cand_party, factor(cand_nm, levels = cand_nm[order(receipt_amt)]))


ggplot(aes(x = cand_nm, y = receipt_amt), data = subset(IL.cand_party))+
  geom_bar(stat = 'identity', position = 'dodge', aes(fill = party))+ 
  coord_flip()
```

From this simple plot only we can identify the fore runners of this election. If
we believe that contributions are a direct reference to the people's choice, then 
we find Hillary Clinton to be leading the race by a great amount. Donald Trump 
comes second and Berie Sander a close third. 
Democratic party seems to be leading with Hillary and Bernie with Republican folllowing
them in the second place. Let us take a look at only the party information, without the 
candidates separated.

```{r, message=FALSE, warning=FALSE}
IL.party <- IL %>%
  group_by(party) %>%
  summarise(receipt_amt = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(party)

ggplot(aes(x = party, y = receipt_amt), data = IL.party)+
  geom_bar(stat = 'identity', aes(fill = party))
```

So, in Illinois, the democrats get more donations than any other party. Almost 2 
times more. And even then, Hillary Clinton gets a major share of the donations t
hat the whole party receives. The independent, libertarian and green party have
such a small amount of contributions, it will be difficult to draw conclusions from
their data. This is one of the reasons that in the later parts of this project
we will be focussing mainly on these two parties(Republican and Democratic) and 
ignoring the rest.

Another idea as to the relative standing of a candidate in his or her own party
is the percentage of contribution they received against the total contributions 
received by the party.

Let us take a quick look at the percentage of donations the members receive.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
# Reshaping IL.cand_party
IL.cand_party.wide <- dcast(IL.cand_party,
                          party ~ cand_nm,
                          value.var = 'receipt_amt')

for(i in 1:nrow(IL.cand_party.wide)){
  IL.cand_party.wide$total[i] <- sum(IL.cand_party.wide[i,2:25], na.rm = TRUE)
}

IL.cand_party.long <- melt(IL.cand_party.wide,
                          id.vars = c('party', 'total'))

IL.cand_party.long <- subset(IL.cand_party.long, !is.na(value))
names(IL.cand_party.long) <- c('party','total','cand_nm','receipt_amt')

IL.cand_party.long$percent <- with(IL.cand_party.long, (receipt_amt*100/total))

# lets look at the percentage of only democrats and republicans
p1 <- ggplot(aes(x = cand_nm, y = percent), 
             data = subset(IL.cand_party.long, party == 'democratic'))+
  geom_histogram(stat = 'identity', aes(fill = I('red4')))+
  ggtitle('Democratic')+
  labs(y = '% of contribution received')+
  theme(axis.text.x = element_text(angle = 90), axis.title = element_text(size = 7),
        axis.title.y = element_text(size = rel(2)), axis.title.x = element_text(size = rel(2)))

p2 <- ggplot(aes(x = cand_nm, y = percent), 
             data = subset(IL.cand_party.long, party == 'republican'))+
  geom_histogram(stat = 'identity', aes(fill = I('dodgerblue4')))+
  ggtitle('Republican')+
  labs(y = '% of contribution received')+
  theme(axis.text.x = element_text(angle = 90), axis.title = element_text(size = 7),
        axis.title.y = element_text(size = rel(2)), axis.title.x = element_text(size = rel(2)))

grid.arrange(p1, p2, ncol = 2)

```


The percentago of donations received tell us about who is leading the financial
race, but it gives us no ideas as to the individual contributions by the contributors.

To get a better sense of the contributions, we could do a box plot. These we will
do one party at a time, since we are interested currently as to how one member stacks 
up against another and what we can deduce from the donations they receive.

### Democratic Party

```{r}
# First lets group the data according to contributors, candidates and the parties
IL.contributor <- IL %>%
  group_by(contbr_nm, cand_nm, party) %>%
  summarise(total_contribution = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(contbr_nm)

# Now lets look at one party at a time
ggplot(aes(x = cand_nm, y = total_contribution), 
       data = subset(IL.contributor, party == 'democratic'))+
         geom_boxplot()+
  scale_y_continuous(limits = c(0,10000), breaks = seq(0,10000,500))

summary(subset(IL.contributor, party == 'democratic'), maxsum = 6)  
```

Hllary Clinton has the largest number of contributors at 19,776. She also has some
of the richest contributors. From the box plot we can see her largest donation 
amount received is a little over $8000. 

Martin O'Malley on the other hand has a larger range of donations and his median
donation received amount is also larger at around $750.

Although the visual range of their donations give us an idea about the the 
distributions, we should not draw conclusions about their total amount received
and such. We should always look at the numbers. A summary of the democratic party 
would  look like the following:
```{r, echo=FALSE}
democratic <- subset(IL, party == 'democratic' ) %>%
  group_by(party, cand_nm) %>%
  summarise(amount = sum(contb_receipt_amt),
            mean = mean(contb_receipt_amt),
            median = median(contb_receipt_amt),
            n = n())
head(democratic)
rm(democratic)
```


### Republican party
```{r, message=FALSE, warning=FALSE}
ggplot(aes(x = cand_nm, y = total_contribution), 
       data = subset(IL.contributor, party == 'republican'))+
         geom_boxplot()+
  scale_y_continuous(limits = c(0,20000), breaks = seq(0,20000,500))+
  theme(axis.text.x = element_text(angle = 90))

summary(subset(IL.contributor, party == 'republican'),maxsum = 16) 
```
The republican party has a lot of members and their distributions are much more
spread out. But again to draw better conclusions we need to look at the numbers.
```{r, echo=FALSE}
republican <- subset(IL, party == 'republican' ) %>%
  group_by(party, cand_nm) %>%
  summarise(amount = sum(contb_receipt_amt),
            mean = mean(contb_receipt_amt),
            median = median(contb_receipt_amt),
            n = n())
head(republican, 16)
rm(republican)
```

One thing is pretty clear: Hillary Clinton is leading the race. But there is a lot 
of similarities between the numbers between Trump and Bernie. It will be interesting 
to see how they compare against each other.

Lets first look at their summaries.

```{r}
candidate <- subset(IL, cand_nm == 'Trump, Donald J.' ) %>%
  group_by(party, cand_nm) %>%
  summarise(amount = sum(contb_receipt_amt),
            mean = mean(contb_receipt_amt),
            median = median(contb_receipt_amt),
            n = n())
head(candidate)

candidate <- subset(IL, cand_nm == 'Sanders, Bernard' ) %>%
  group_by(party, cand_nm) %>%
  summarise(amount = sum(contb_receipt_amt),
            mean = mean(contb_receipt_amt),
            median = median(contb_receipt_amt),
            n = n())
head(candidate)
```
```{r, echo=FALSE}
rm(candidate)
```


Ok, so the total amounts look close enough. But that does not give us a clear idea
as to how their contributions are distributed. For a visual comparison lets plot 
a freqpoly of the two candidates on the amount that they cumulatively received.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# A grouping to be primarily used in the later stage, but brought forward for this plot
IL.party_date <- IL %>%
  group_by(contb_receipt_dt, cand_nm, party) %>%
  summarise(total_amnt = sum(contb_receipt_amt),
            mean_amnt = mean(contb_receipt_amt),
            median_amnt = median(contb_receipt_amt),
            n = n()) %>%
  arrange(contb_receipt_dt)

# Freqpoly
ggplot(aes(x = total_amnt), 
       data = subset(IL.party_date, cand_nm %in% c('Trump, Donald J.','Sanders, Bernard')))+
  geom_freqpoly(aes(color = cand_nm))+
  scale_colour_brewer(type = 'qual', guide = guide_legend(title = 'candidate', reverse = F))+
  xlim(c(0,NA))+
  ylim(c(0,NA))
```

### CONTRIBUTORS

We start off by plotting histogram of the number of donors by the amount they 
donated.
```{r}
# what amount does the donors donate. Summary of doation
summary(IL.contributor$total_contribution)

ggplot(aes(x = total_contribution), data = IL.contributor)+
  geom_histogram(binwidth = 10, color = I('black'), aes(fill = I('Orange')))+
  scale_x_continuous(limits = c(0,1500), breaks = seq(0, 1500, 50))
```

And as expectd, we find a larger number of people in the lower range. Also 
noticeable are the peaks at certain figures like 250, 500, 600 and so on. These  
are also in the bounds fo our expectations. People are more likely to donate $600
than say 610 or 590. Hence the peaks.

### SCATTER PLOT (contributors vs contribution)
Another perspective of the same data may be observed by plotting a scatter plot
of the individual contributors vs amount donated and then classifying them with 
the party they contributed to.
```{r}
IL.contributor$party <- with(IL.contributor, factor(party, 
                        levels = c('democratic','republican','libertarian',
                                       'green','independent')))
#str(IL.contributor$party)

ggplot(aes(x = total_contribution, y = n, color = party), data = IL.contributor)+
  geom_point(alpha = 0.4, position = 'jitter')+
  xlim(0, quantile(IL.contributor$total_contribution,0.9))+
  ylim(0, quantile(IL.contributor$n,0.9))+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'party', reverse = F))
```

This graph gives some interesting results. 

First, we see the absolute disparity between the democratic and republican parties
with respect to the other parties. The number of individuals contributing to the 
3 lesser known political parties are almost insignificant to the leading ones.

Corresponding to the peaks we found in the earlier histograph, we also find certain 
parallel lines formed by the dots that are formed due to more number of cotributors
at certain amounts.


### MEDIAN TREND

Although illuminating, the above graphs dont really show us a trend of the contributors
with respect to the amount they are willing to contribute to their parties.
An interesting plot from this perspective would be to look at the median amounts
received by each party by the number of people donating these median amounts.
```{r, message=FALSE, warning=FALSE}
ggplot(aes(x = median_amnt), data = IL.party_date)+
  geom_freqpoly(aes(color = party), na.rm = TRUE)+
  scale_colour_brewer(type = 'qual', guide = guide_legend(title = 'party', reverse = F))+
  xlim(c(0,NA))+
  ylim(c(0,175))
```

### TIME FRAME

An important feature that we have neglected to explore all this time has been the 
time data. There are four years between elections, and the donations happen over 
time throughout these four years.

It will be interesting to note how the contributions vary over time for the 
individual parties.
```{r}
ggplot(aes(x = contb_receipt_dt, y = total_amnt, color = party), 
       data = IL.party_date)+
  geom_line(na.rm = TRUE)+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'party', reverse = F))+
  scale_x_date(limits = c(dmy('01-01-2015'), dmy('01-01-2017')))+
  ylim(c(0,NA))
```

This plot is not gradual in its trend at all. It is mostly comprised of sudden 
peaks of donations and then gaps of nil or very small donations. This is interesting.
The reason, without going into more details or data hunting on the web, may be 
that these donations correspond with certain political events or decisions by the 
parties that influenced the supporters to show their support in the form of
donations.

The trend of the donations increasing towards the end is very logical and expected.
It is obvious that more people tend to donate as the presidential election date nears
and they have had a chance to make up their mind as to who they want to support.

We can do the same thing for individual candidates that we have done above for
the political parties as a whole. 

Here we only take a closer look at the democratic and republican parties. 
```{r}
p1 <- ggplot(aes(x = contb_receipt_dt, y = total_amnt, color = cand_nm), 
       data = subset(IL.party_date, party == 'democratic'))+
  geom_line(na.rm = TRUE)+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'candidate', reverse = F))+
  scale_x_date(limits = c(dmy('01-01-2015'), dmy('01-01-2017')))+
  ylim(c(0,NA))

exclude = c("Jindal, Bobby", "Santorum, Richard J.", "Huckabee, Mike", 
            "Perry, James R. (Rick)", "Pataki, George E.")

p2 <- ggplot(aes(x = contb_receipt_dt, y = total_amnt, color = cand_nm), 
       data = subset(IL.party_date, (party == 'republican' & cand_nm %!in% exclude)))+
  geom_line(na.rm = TRUE)+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'candidate', reverse = F))+
  scale_x_date(limits = c(dmy('01-01-2015'), dmy('01-01-2017')))+
  ylim(c(0,NA))

grid.arrange(p1, p2, ncol = 1)
```


### GEOGRAPHICAL ASPECTS

We know about the zip codes of the contributors. So, it would be interesting to 
see people from which places are contributing and who are they contributing to.
We could plot the different locations that people are contributung from on a map.

If we are lucky, and there are some places which are overwhelmingly in favour of 
one party or the other, we would see clear distinctions of that as well.

For starters, let us look at the map of USA populated by the donors according to 
their zip codes converted to longitudes and latitudes.
```{r, echo=FALSE}
# Read a new file that contained zip with corresponding longitudes and latitudes

## The following code to get the longitudes and latitudes take a long time. So I have done the 
## conversion once and saved the result in a file for easy access the next time I run the code

#zip <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/zip_codes_states.csv')

#IL.zip <- IL[c(3,5,7,10,11,19)]
#IL.zip$long = NA
#IL.zip$lat = NA

# inserting long & lat
#for(i in 1:length(IL.zip$contbr_zip)){
#   indices <- which(zip$zip_code == IL.zip$contbr_zip[i])
#   IL.zip$lat[i] <- ifelse(length(indices)>0,zip$latitude[indices],NA)
#   IL.zip$long[i] <- ifelse(length(indices)>0,zip$longitude[indices],NA)
#}

#write.csv(IL.zip, file = "IL_zip.csv")

IL.zip <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/IL_zip.csv')
```

```{r}
all_states <- map_data("state")

# state map with dots
ggplot(aes(x = long, y = lat), data = IL.zip) +
  geom_polygon(data = all_states, aes(x=long, y=lat, group = group),colour="white", fill="grey10")+
  geom_point(color = "dark red", na.rm = TRUE)+
  coord_fixed(1.3)
```

So, for the state of illinois, ofcourse the donors will also be located in the 
same state. But we do find some donors outside of the state lines. 

Could they be outliers? Are they informations wrongly entered into they system?
Could be. Some other explanations may be people who were originally from Illinois,
or still are and are temporarily located else where are probably donating money 
from their temporary locations. Maybe some organisations are donating money for 
to local candidate, but from institutions that hold their money and are located 
outside the state. 

In any case, with no further information, it is difficult for us to come to any
conclusion. In the mean time, let us overlook these locations that are outside the 
state borders and take a closer look at all of the contributions from within state.


```{r}
# only chicago with dots
county <- map_data('county')

ggplot(aes(x = long, y = lat), data = subset(IL.zip, long >= '-87.49638' & long <= '-91.50136'
                                             & lat >= '37.00161' & lat <= '42.50774')) +
  geom_polygon(data = subset(county, region == 'illinois'), 
               aes(x=long, y=lat, group = group),colour="white", fill="grey10")+
  geom_point(color = "dark red", na.rm = TRUE, alpha = 0.4, position = 'jitter')+
  coord_fixed(1.3)
```

The distribution of donors do not show much character to our dissapointment. They 
more or less evenly distributed. What remains to be seen is how the distribution 
would vary once we add in the amount that each of these places contributed.

### CONTRIBUTIONS SCALED BY DOT SIZE

```{r}
# scaling dot size
IL.zip_grouped <- subset(IL.zip, contb_receipt_amt > 0)

IL.zip_grouped$amnt[IL.zip_grouped$contb_receipt_amt >= 80] <- '10800'
IL.zip_grouped$amnt[IL.zip_grouped$contb_receipt_amt >= 27 & 
                      IL.zip_grouped$contb_receipt_amt <= 80] <- '80'
IL.zip_grouped$amnt[IL.zip_grouped$contb_receipt_amt >= 15 & 
                      IL.zip_grouped$contb_receipt_amt <= 27] <- '27'
IL.zip_grouped$amnt[IL.zip_grouped$contb_receipt_amt <= 15] <- '15'

IL.zip_grouped$amnt <- as.numeric(IL.zip_grouped$amnt)

ggplot(aes(x = long, y = lat), data = subset(IL.zip_grouped, long >= '-87.49638' & long <= '-91.50136'
                        & lat >= '37.00161' & lat <= '42.50774' & contb_receipt_amt > 0)) +
  geom_polygon(data = subset(all_states, region == 'illinois'), 
               aes(x=long, y=lat, group = group),colour="white", fill="grey10")+
  geom_point(na.rm = TRUE, alpha = 0.5, position = 'jitter',
             aes(color = party, size = amnt))+
  scale_size_continuous(range=c(0.5,4.5), name = 'Amount')+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'party', reverse = F))+
  coord_fixed(1.3)
```

This gives us a better idea as to which locations contributed more money and 
those that had little to contribute. There are some places that are completely blank.
These could be places with poor economic standing or less population density.

### CONTRIBUTIONS BY COUNTY
Now, the above visualization has considered contributions from particular 
locations, as in zip codes to be exact. We can group these and view them as 
contributions from different counties as well. 

```{r, echo=FALSE}
county <- map_data('county')

IL.county_grouped <- IL.zip_grouped %>%
  group_by(contbr_zip, party) %>%
  summarise(amount = sum(contb_receipt_amt),
            n = n())

county_zip <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/county_zip.csv')
county_population <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/illinois_county_population2.csv')
county_zip$county <- as.character(county_zip$county)

IL.county_grouped$county = ''
for(i in 1:length(IL.county_grouped$contbr_zip)){
  index <- which(county_zip == IL.county_grouped$contbr_zip[i])
  IL.county_grouped$county[i] <- ifelse(length(index)>0, county_zip$county[index], 0)
}

IL.county_cum <- subset(IL.county_grouped, !is.na(county)) %>%
  group_by(county) %>%
  summarise( amount = sum(amount))

sub_county <- subset(county, region == 'illinois')
#sub_county <- sub_county[c(1,2,3,5,6)]

sub_county$county_name <- paste0(sub_county$subregion,' County')

sub_county$contribution <- NA
sub_county$population <- NA
for(i in 1:length(sub_county$long)){
  index <- which(toupper(sub_county$county_name[i]) == toupper(IL.county_cum$county))
  ifelse(length(index)>0,sub_county$contribution[i] <- IL.county_cum$amount[index],0)
  index <- which(toupper(sub_county$county_name[i]) == toupper(county_population$County))
  ifelse(length(index)>0,sub_county$population[i] <- county_population$Population[index],0)
}

sub_county$cont_by_popl = with(sub_county, contribution/population)

ggplot(data = sub_county) + 
  geom_polygon(aes(x = long, y = lat, fill = contribution, group = group), color = "black")+ 
  coord_fixed(1.3)+
  scale_fill_gradient(trans = "log10")
```

But there is no relativity here. Only a difference by which the total contributions 
have varied from one county to the other.

To add the population factor here, we need to look at the ratio of contribution to
population for each country. 

### Ratio of Contribution to population per county

```{r}
ggplot(data = sub_county) + 
  geom_polygon(aes(x = long, y = lat, fill = cont_by_popl, group = group), color = "black")+ 
  coord_fixed(1.3)+
  scale_fill_gradient(trans = "log10")
```


### A time frame added to geographical plot

We have already seen one time frame plot. We now add those data to the map above,
and look how people from different locations have donated over time from a 
geographical point of view.

```{r}
IL.zip_grouped$date <- with(IL.zip_grouped, as.character(year(contb_receipt_dt)))

colors <- c("gold",
            "red4",
            "cornsilk",
            "darkgreen")

ggplot(aes(x = long, y = lat), data = subset(IL.zip_grouped, 
                                  long >= '-87.49638' & long <= '-91.50136'
                                  & lat >= '37.00161' & lat <= '42.50774' 
                                  & contb_receipt_amt > 0)) +
  geom_polygon(data = subset(all_states, region == 'illinois'), 
               aes(x=long, y=lat, group = group),colour="white", fill="grey10")+
  geom_point(na.rm = TRUE, alpha = 0.5, position = 'jitter',
             aes(color = date), shape = 21)+
  scale_color_manual(values = colors)+
  coord_fixed(1.3)
```

### CONTRIBUTORS BY POPULATION

We have looked at the ratio of contribution to population from a geographical
perspective. But contributions of an individual is dependent upon a lot of factors,
primary of which is his economic standing.

The ratio of amount divided by population does not necessarily portray a direct
correlation with the number of people supporting a candidate. For this we need
to look not at the amount of contribution, but at the number of contributors
with respect to the population.

```{r, echo=FALSE, message = FALSE, warning=FALSE}
IL.county_grouped <- IL.zip_grouped %>%
  group_by(contbr_zip, party) %>%
  summarise(amount = sum(contb_receipt_amt),
            n = n())

county_zip <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/county_zip.csv')  
county_zip$county <- as.character(county_zip$county)

IL.county_grouped$county <- ""
for(i in 1:length(IL.county_grouped$contbr_zip)){
  index <- which(county_zip$zip == IL.county_grouped$contbr_zip[i])
  IL.county_grouped$county[i] <- ifelse(length(index)>0, county_zip$county[index], NA)
}

county_zip.county <- IL.county_grouped %>%
  group_by(county, party) %>%
  summarise(amnt = sum(amount),
            num = sum(n))


county_population <- read.csv('/Users/sohansamanta/R/Data Analysis with R/PROJECT/illinois_county_population2.csv')
county_zip.county$population <- ""

for(i in 1:length(county_zip.county$county)){
  index <- which(county_population$County == county_zip.county$county[i])
  county_zip.county$population[i] <- ifelse(length(index)>0, county_population$Population[index], NA)
}

county_zip.county$population <- as.integer(county_zip.county$population)

ggplot(aes(x = population, y = num, color = party), data = county_zip.county)+
  geom_smooth(na.rm = TRUE)+
  scale_x_log10()+
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'candidate', reverse = F))

```

No surprise here. From our earlier explorations, we expected a result close to this.

### CONTRIBUTIONS BY OCCUPATION

A last part that I would like to touch before wrapping this up is the occupation
of the contributors.The total variation of occupations are 8364, we wont look at
all of them.

Instead we would select a few occupations that have the most number of contributors 
and study who they contributed towards.
```{r}
length(unique(IL$contbr_occupation))

occupation <- IL %>%
  group_by(contbr_occupation, party) %>%
  summarise(amount = sum(contb_receipt_amt),
            n=n()) %>%
  arrange(n)


occp = c('RETIRED','NOT EMPLOYED','ATTORNEY','TEACHER', 'PROFESSOR', 'PHYSICIAN', 
         'CONSULTANT','HOMEMAKER', 'LAWYER', 'SALES', 'STUDENT', 'ENGINEER')


ggplot(aes(x = party, y = amount), data = subset(occupation, contbr_occupation %in% occp))+
         geom_bar(stat = 'identity', aes(fill = party))+
         facet_wrap(~contbr_occupation)+
  theme(axis.text.x = element_text(angle = 45))
```

-------------------------------------------------------------------------------
-------------------------------------------------------------------------------


# Final Plots and Summary

This data set is essentially a log of how much money was contributed by individuals
and institutions to the candidates from different political parties from the state
of illinois who were contesting in the Presidential Election of 2016.

One of the initial plots which according to me was also one of the most significant 
ones is the one given below:

### Plot One
```{r echo=FALSE, Plot_One}
p1 <- ggplot(aes(x = cand_nm, y = receipt_amt), data = subset(IL.cand_party))+
  geom_bar(stat = 'identity', position = 'dodge', aes(fill = party))+ 
  coord_flip()+
  labs(y='Amount Received', title = 'Cumulative contributions to candidates',
  x='Candidate Names')
  
p2 <- ggplot(aes(x = party, y = receipt_amt), data = IL.party)+
  geom_bar(stat = 'identity', aes(fill = I('dark red')))+
  coord_flip()+labs(y='Amount Received', x='Party Names')
  
grid.arrange(p1, p2, ncol = 1)
```

### Description One
The immense disparity between the amounts donated to each of the parties and the 
political candidate who is the sole reason for the this huge margin is clear from 
the above graphs. Hillary Clinton not only commands a little above 80% of the 
total fund collected by the democratic party, she also towers amongst her contestants
by leaving behind her close second Donald Trump by almost 12 million dollars.

This difference is significant, and gives a clear view of the political favourite
from the state of illinois, at least, as far as donating citizens are concerned.


### Plot Two
```{r echo=FALSE, Plot_Two}
candidates = c('Clinton, Hillary Rodham', 'Trump, Donald J.', 'Sanders, Bernard',
               'Rubio, Marco', 'Bush, Jeb', "Cruz, Rafael Edward 'Ted'")

p1 <- ggplot(aes(x = median_amnt), data = subset(IL.party_date, cand_nm %in% candidates))+
  geom_freqpoly(aes(color = cand_nm), na.rm = TRUE, binwidth = 5)+
  scale_colour_brewer(type = 'qual', guide = guide_legend(title = 'candidates', reverse = F))+
  xlim(c(0,500))+
  ylim(c(0,NA))+
  labs(x='Median Amount', title = 'Frequency polygraph of No. of candidates per mean amount',
       y='Number of candidates')

p2 <- ggplot(aes(x = median_amnt), data = subset(IL.party_date, cand_nm %in% candidates &
                                             median_amnt > 0))+
  geom_freqpoly(aes(color = cand_nm), na.rm = TRUE, binwidth = 5)+
  scale_colour_brewer(type = 'qual', guide = guide_legend(title = 'candidates', reverse = F))+
  #xlim(c(0,NA))+
  #ylim(c(0,125))+
  labs(x='Median Amount(log scale)', title = 'Frequency polygraph of No. of candidates per mean amount',
       y='Number of candidates')+
  scale_x_sqrt()

grid.arrange(p1,p2,ncol=1)
```

### Description Two
This is an interesting plot and gives us a good idea of the variation of the 
contributions. It is interesting to note that the median amount of donations for
three candidates, Clinton, Cruz and Sanders have peaks at the same median. This tells
us that the median amount of contributors for all three of these candidates probably
have comparable economic standings. Trump however has a median amount a little higher 
than them. It can also be seen that both Jeb and Rubio have peaks at higher median
values, but their peaks are so small, the number of people corresponding to them
are quite frankly not of much significance on the grand scale of things.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = sub_county) + 
  geom_polygon(aes(x = long, y = lat, fill = cont_by_popl, group = group), color = "black")+ 
  coord_fixed(1.3)+
  scale_fill_gradient(trans = "log10")+
  labs(title = 'Ratio of contribution to population per county')
```

### Description Three
This color scale across counties representing the contributions per county divided
by the population of that county is interesting. We see a very clear picture of
which counties contributed more per person and which did not. Most significant are 
grey counties which represent no contributions what so ever. Now this could be 
considered as outliers caused by missing zip codes for contributors, but even then
this is too large a discrepancy to be a coincidence. That leaves the fact that either
the contribution for these regions are too low or the population is so high so as
to decrease the small contributions negligible.
------

# Reflection

I approached this project by first stating a few questions that I felt one would like 
to know from a financial data set relating to a presidential election. As I thought 
about answering these questions, I thought about the information required and the 
information available in the data set. I then proceeded to answer them by first 
looking at the data structure and summary and then selecting variables from these 
summaries and then working on them. 
After each plot and analysis, I revisited my questions and came up with new ones.
My approach was to go over the data and find correlations between the variables
I was working with. Once a relation was established, I would dig deeper to see if
more inferences could be drawn by maybe adding new data or sometimes even removing 
some data.
It took me more time to complete than I had initally expected. While working I found 
myself to be continually visiting my previous works and changing them to make minute
adjustments based on some new information uncovered at a later stage.


### Sources
1. Main dataset from the Federal Election Commission 
2. Stack overflow for ideas
3. Population by county from web
4. Zip code by county from web